### Week4
# Workthrough
# Introduction to tidyverse


## Copy files from CytometryInR into week4 folder

```{r}
getwd()
```


```{r}
Sourcepath <- "~/Documents/CytometryInR/course/04_IntroToTidyverse/data/"
targetpath <- "~/Documents/CytometryInR course/Week4/data/"
files_newweek4 <- list.files(Sourcepath, full.names = TRUE, recursive = TRUE)
file.copy(from = files_newweek4, to=targetpath)
```

## read.csv
We will start by first loading in our copied over dataset (Dataset.csv) from it’s location in the project folder. 


```{r}
thefilepath <- file.path("data", "Dataset.csv")
thefilepath
```

Above, we directly specified the name (Dataset) and filetype (.csv) of the file we wanted in the last argument of the file.path (“Dataset.csv”). This allows us to skip the list.files() step we used last week as we have provided the full file path. 

Since our dataset is stored as a .csv file, we will be using the read.csv() function from the utils package (included in our base R software installation) to read it into R. We will also use the colnames() function from last week to get a read-out of the column names.

```{r}
Data <- read.csv(file=thefilepath, check.names=FALSE)
colnames(Data)
```

As we look at the line of code, we now have enough context to decipher that the “file” argument is where we provide a file path to an individual file, but what does the “check.names” argument do?

Let’s see what happens to the column names when we set “check.names” argument to TRUE:
```{r}
Data_Alternative <- read.csv(file=thefilepath, check.names=TRUE)
colnames(Data_Alternative)
```

As we can see, any column name that contained a special character or a space was automatically converted over to R-approved syntax. However, this resulted in the loss of both +” and “-”, leaving us unable to determine whether we are looking at cells within or outside a particular gate.

```{r}
head(Data, 3)
```

We could also achieve the same window to open using the View() function:
```{r}
View(Data)
```

Wrapping up our brief recap of last week functions, we can check an objects type using both the class() and str() functions.

```{r}
class(Data)
```

```{r}
str(Data)
```

# data.frame
Or alternatively using the new-to-us glimpse() function

```{r}
glimpse (Data)
```

Error in `glimpse()`:
! could not find function "glimpse"
 

 ```{r}
 ??glimpse
 ```

 From the list of search matches (in the right secondary sidebar), it looks likely that the glimpse() function in the dplyr package was the one we were looking for. This is one the main tidyverse packages we will be using throughout the course. Let’s attach it to our environment via the library() call first and try running glimpse() again.

```{r}
library(dplyr)
```

```{r}
glimpse(Data)
```

We notice that while similar to the str() output, glimpse() handles spacing a little differently, and includes the dimensions at the top. However, we can also retrieve the dimensions directly using the dim() function (which maintains the row followed by column position convention of base R (ex. [196,31]))


```{r}
dim(Data)
```

## Column value type

If we were trying to verify type of values contained within a data.frame column, we could employ several similarly-named functions (is.character(), is.numeric() or is.logical()) to check

```{r}
is.character(Data$bid)
```

```{r}
is.numeric(Data$bid)
```

```{r}
is.character(Data$Tcells_count)
```

For numeric columns using the is.numeric() function, we can also be subtype specific using either is.integer() or is.double().

```{r}
is.numeric(Data$Tcells_count)
```

```{r}
is.integer(Data$Tcells_count)
```
```{r}
is.double(Data$Tcells_count)
```

# NB: As we observed last week with keywords, column names that contain special characters like $ or spaces will need to be surrounded with tick marks in order for the function to be able to run.

```{r}
is.numeric(Data$CD8-)
```

Error in parse(text = input): <text>:2:21: unexpected ')'
1: # colnames(Data)  # To recheck the column names
2: is.numeric(Data$CD8-)

```{r}
is.numeric(Data$"CD8-")
```

# select (Columns)
Now that we have read in our data, and have a general picture of the structure and contents, lets start learning the main dplyr functions we will be using throughout the course. To do this, lets go ahead and attach dplyr to our local environment via the library() call.

```{r}
library(dplyr)
```

We will start with the select() function. It is used to “select” a column from a data.frame type object. In the simplest usage, we provide the name of our data.frame variable/object as the first argument after the opening parenthesis. This is then followed by the name of the column we want to select as the second argument (let’s place around the “” around the column name for now)

```{r}
DateColumn <- select(Data, "Date") 
DateColumn[1:10,]
```

This results in the column being selected, resulting in the new object containing only that subsetted out column from the original Data object.

## Pipe Operators
While the above line of code works to select a column, when you encounter select() out in the wild, it will more often be in a line of code that looks like this:

```{r}
DateColumn <- Data |> select("Date")
DateColumn [1:10,]
```

… “What in the world is that thing |> ?” …

Glad you asked! An useful feature of the tidyverse packages is their use of pipes (either the original magrittr package’s “%>%” or base R version >4.1.0's “|>”“), usually appearing like this:

```{r}
# magrittr %>% pipe

DateColumn <- Data %>% select("Date")

# base R |> pipe
DateColumn <- Data |> select("Date")
DateColumn
```

Let’s break it down, starting off just to the right of the assignment arrow (<-) with our data.frame “Data”.

```{r}
Data
```

We then proceed to read to the right, adding in our pipe operator. The pipe essentially serves as an intermediate passing the contents of data onward to the subsequent function.

```{r}
Data |> select("Date")
```

In summary, contents of Data are passed to the pipe, and select runs on those contents to select the Date column

One of the main advantages for using pipes, is they can be linked together, passing resulting objects of one operation on to the next pipe and subsequent function. We can see this in operation in the example below where we hand off the isolated “Date” column to the nrow() function to determine number of rows. We will use pipes throughout the course, so you will gradually gain familiarity as you encounter them.

```{r}
Data |> select("Date") |> nrow()
```

## R Quirks

Odd R Behavior # 1
While we used “” around the column name in our previous example, unlike what we encountered with install.packages() when we forget to include quotation marks, select() still retrieves the correct column despite Date not being an environment variable:

```{r}
Data |> select(Date) |> head(3)
```

## Selecting multiple columns

Since we are able to select one column, can we select multiple (similar to a [Data[,2:5]] approach in base R)? We can, and they can be positioned anywhere within the data.frame:


```{r}
Subset <- Data |> select(bid, timepoint, Condition, Tcells, 'CD8+', 'CD4+')

head(Subset, 3)
```


You will notice that the order in which we selected the columns will dictate their position in the subsetted data.frame object:

```{r}
Subset <- Data |> select(bid, Tcells, `CD8+`, `CD4+`, timepoint, Condition, )

head(Subset, 3)
```

## relocate
Alternatively, we occasionally want to move one column. While we could respecify the location using select(), specifying the names of all the other columns out in a line of code to just to rearrange one does not sound like a good use of time. For this reason, the second dplyr function we will be learning is the relocate() function.

Looking at our Data object, let’s say we wanted to move the Tcells column from its current location to the second column position (right after the bid column). The line of code to do so would look like:

```{r}
Data |> relocate(Tcells, .after=bid) |> head(3)
```

# |> head(3) is used only to make the website output visualization manageable :D

Similar to what we saw with select(), this approach can also be used for more than 1 column:

```{r}
Data |> relocate(Tcells, Monocytes, .after=bid) |> head(3)
```

We can also modify the argument so that columns are placed before a certain column

```{r}
Data |> relocate(Tcells, .before=Date) |> head(3)
```

And as we might suspect, we could specify a column index location rather than using a column name.

```{r}
Data |> relocate(Date, .before=1) |> head(3)
```

## rename

How can we change a column name?

In base R, we could change individual column names by assigning a new value with the assignment arrow to the corresponding column name index. For example, looking at our Subset object, wen could rename CD8+ as follows:


```{r}
colnames(Subset)
```

```{r}
colnames(Subset)[3]
```

```{r}
colnames(Subset) [3] <- "CD8Positive"

colnames(Subset)
```

With the tidyverse, we can use the rename() function which removes the need to look up the column index number. The way we write the argument is placing within the parenthesis the old name to the right of the equals sign, with the new name to the left


```{r}
renamed <- Subset |> rename(CD4_Positive = "CD4+")

colnames(renamed)
```

If we wanted to rename multiple column names at once, we would just need to include a comma between the individual rename arguments within the parenthesis.

```{r}
Renamed_Multiple <- Subset |> rename (specimen = bid, timepoint_months = timepoint, stimulation = Condition, CD4Positive="CD4+")

colnames(Renamed_Multiple)
```

## Pull

Sometimes, we may want to retrieve individual values present in a column, to use within either a vector or a list. We can do this using the pull() function, which will retrieve the column contents and strip the column formatting

```{r}
Data |> pull(Date) |> head(5)
```

This can be useful when we are doing data exploration, and trying to determine how many unique variants might be present. For example, if we wanted to see what days individual samples were acquired, we could pull() the data and pass it to the unique() function:

```{r}
Data |> pull(Date) |> unique()
```

## filter (Rows)

So far, we have been working with dplyr functions primarily used when working with and subsetting columns (including select(), pull(), rename() and relocate()). What if we wanted to work with rows of a data.frame? This is where the filter() function is used.

The Condition column in this Dataset appears to be indicating whether the samples were stimulated. Let’s see how many unique values are contained within that column

```{r}
Data |> pull(Condition) |> unique()
```

In the case of this dataset, looks like the .fcs files where treated with either left alone, treated with PPD (Purified Protein Derrivative) or SEB. What if we wanted to subset only those treated with PPD?

Within filter(), we would specify the column name as the first argument, and ask that only values equal to (==) “PPD” be returned. Notice in this case, “” are needed, as we are asking for a matching character value.

```{r}
PPDOnly <- Data |> filter(Condition == "PPD")

head(PPDOnly, 5)
```

While this works, using “==” to match can glitch, especially with character values. Using the %in% operator is a better way of identifying and extracting only the rows whose Condition column contains “PPD”

```{r}
Data |> filter(Condition %in% "PPD") |> head(5)
```

Similar to what we saw for select(), we can grab rows that contain various values at once. We would just need to modify the second part of the argument. If we wanted to grab rows whose Condition column contained either PPD or SEB, we would need to provide that argument as a vector, placing both within c()/

```{r}
Data |> filter(Condition %in% c("PPD", "SEB")) |> head(5)
```

Alternatively, we could have set up the vector externally, and then provided it to filter()

```{r}
TheseConditions <- c("PPD", "SEB")
Data |> filter(Condition %in% TheseConditions) |> head(5)
```

While this works when we have a limited number of variant condition values, what if had many more but only wanted to exclude one value? As we saw when learning about Conditionals, when we add a ! in front of a logical value, we get the opposite logical value returned

```{r}
IsThisASpectralInstrument <- TRUE

!IsThisASpectralInstrument
```

In the context of the dplyr package, we can use ! within the filter() to remove rows that contain a certain value

```{r}
Subset <- Data |> filter(!Condition %in% "SEB")
Subset |> pull(Condition) |> unique()
```

Likewise, we can also use it with the select() to exclude columns we don’t want to include

```{r}
Subset <- Data |> select(!timepoint)
Subset[1:3,]
```

## mutate
But what if we wanted to alter the content of a column, or add new columns to an existing data.frame? This is where the mutate() function can be used.

Let’s start by slimming down our current Data to a smaller workable example, highlighting the functions and pipes we learned about today

```{r}
TidyData <- Data |> filter(Condition %in% "Ctrl") |> filter(timepoint %in% "0") |> select(bid, timepoint, Condition, Date, Tcells_count, CD45_count) |> rename(specimen=bid, condition=Condition) |> relocate (Date, .after=specimen)
```

```{r}
TidyData
```

The mutate() function can be used to modify existing columns, as well as to create new ones. For example, let’s derrive the proportion of T cells from the overall CD45 gate. To do so, within the parenthesis, we would specify a new column name, and then divide the original columns:


```{r}
TidyData <- TidyData |> mutate(Tcells_ProportionCD45 = Tcells_count / CD45_count)
TidyData
```

We can see that we have many significant digits being returned. Let’s round this new column to 2 significant digits by applying the round() function


```{r}
TidyData <- TidyData |> mutate(TcellsRounded = round(Tcells_ProportionCD45, 2))
TidyData 
```

## arrange

And while we are here, let’s rearrange the rows so that they are descending based on the Tcell proportion. We can use this by using the desc() and arrange() functions from dplyr:

```{r}
TidyData <- TidyData |> arrange(desc(TcellsRounded))
```

And let’s go ahead and filter() and identify the specimens that had more than 30% T cells as part of the overall CD45 gate (context, these samples were Cord Blood):

```{r}
TidyData |> filter(TcellsRounded > 0.3)
```

Which is we had wanted to just retrieve the specimen IDs, we could add pull() after a new pipe argument.

```{r}
TidyData |> filter(TcellsRounded > 0.3) |> pull(specimen)
```

And finally, since I may want to send the data to a supervisor, let’s go ahead and export this “tidyed” version of our data.frame out to it’s own .csv file. Working within our project folder, this would look like this:

```{r}
NewName <- paste0("MyNewDataset", ".csv")
StorageLocation <- file.path("data", NewName)
StorageLocation
```


```{r}
write.csv(TidyData, StorageLocation, row.names=FALSE)
```
